#RStudio Exercise 4 - Clustering and classification

```{r MASS}
library(MASS)
data("Boston")
```

There are many data sets already loaded in R or included in the package installations. In this exercise we will practise our analysis on a data set called "Boston" from the "MASS" package. 
The data set concernes the housing values in the suburbs of Boston.

```{r head.Boston}
head(Boston)
```

```{r dim Boston}
dim(Boston)
str(Boston)
```

```{r summary.variables}
summary(Boston)
```

```{r pairs.boston}
pairs(Boston)
```
```{r boston.atrix}
library(tidyr)
library(corrplot)
cor_matrix<-cor(Boston)
cor_matrix %>% round(digits=2)
corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
```

####Data wrangling 
Standardizing the dataset and creating a factor variable for the crime rate. As the data contains only numerical values, we can use the scale() function to standardize the whole dataset. Dividing the dataset into a train and a test set.

```{r boston.scaled}
boston_scaled <- scale(Boston)
summary(boston_scaled)
```
 The variable crim describes the per capita crime rate by town.
```{r variable}
class(boston_scaled)
boston_scaled <- as.data.frame(boston_scaled)
bins <- quantile(boston_scaled$crim)
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels = c("low", "med_low", "med_high", "high"))
boston_scaled <- dplyr::select(boston_scaled, -crim)
boston_scaled <- data.frame(boston_scaled, crime)
head(boston_scaled)
```

```{r split}
n <- nrow(boston_scaled)
ind <- sample(n, size = n*0.8)
train <- boston_scaled[ind,]
test <- boston_scaled[-ind,]
correct_classes <- test$crime
test <- dplyr::select(test, -crime)
dim(test)
dim(train)
```

Fitting a linear discriminant analysis to the training set. Crime is a target varaible and all the other variables are predictors.

```{r discr.analysis}
lda.fit <- lda(crime ~ ., data = train)
lda.fit
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}
classes <- as.numeric(train$crime)
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

```

```{r predict.crime}
lda.pred <- predict(lda.fit, newdata = test)
table(correct = correct_classes, predicted = lda.pred$class)
```